Product plots

### 2011-06-27 reviews/response.txt

We are grateful for the helpful feedback provided by the reviewers. We have added a related work section, including all of the references we missed. We have also weakened our claims about the superiority of rectangles, instead providing a more pragmatic justification

@@

Other changes:

* Rather than eliminating the section describing the R implementation altogether, we have expanded it slightly to give a taste of how it might be used.

* Figures are labelled with distributional decomposition.

* We have expanded the section that introduces joint, marginal and conditional densities to be more helpful for newcomers to the topic.

@@

* We wish we could make the figures larger and the labels easier to read, but the page limits are extremely tight this year.  We have done our best to ensure that all information is convey redundantly through colour or description in the captions.

* We have endeavoured to make all of the other minor improvements to spelling and grammar that the reviewers suggested.  Thanks!

### 2011-06-27 prodplots-infovis.tex
Final tweaks

We begin in Section~\ref{sec:primitives} by motivating the three constraints that give rise to the basic graphical component of product plots: the rectangle. We constrain the rectangle further to produce three 1d atoms (bar, spine, tile) and one 2d atom (the fluct). These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include more visualisations such as histograms, pie charts, cascaded treemaps and weighted plots.

@@

The product plots framework is heavily influenced by the work of Wilkinson \citep{wilkinson:1999, wilkinson:2005}, who suggested that both mosaic plots and treemaps can be described as plots with coordinate systems based on recursive partitioning. Our work builds on this by adding conditioning, and defining a ``grammar of area plots'', a domain specific language for describing, in detail, this small family of plots. Other similar efforts to build domain specific languages for visualisation are APT \citep{mackinlay:1986} and VisQL \citep{mackinlay:2007}: our framework is driven by similar forces, but carves out a much smaller niche. This makes products plots less expressive, but more efficient, because the number of primitives required to describe the smaller domain is much smaller.

@@

To construct plots of higher-dimensional data sets, we need a way to decompose them into 1d and 2d components. Some statistical vocabulary is useful: one way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}. A {\sc pmf} is a function with $n$ inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs. A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all values must sum to one.

@@

Figure~\ref{fig:2d-table} shows three ways to represent the the 2d table of proportions of sex and happiness. The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' -- the bottom left number tells us that these are 0.14. The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex. These correspond to restricting the row or column sums to one, and support questions such as ``what proportion of pretty happy people are female?'' --  the table tells us that it's 0.55, or ``what proportion of males are not too happy?'' -- the top left number in the table tells us that that's 0.12. Finally, the last table displays the two {\bf marginal distributions} of sex and happiness. These allow us to answer questions like ``what proportion of respondents were male'' (0.44) or ``what proportion of respondents were very happy?'' (0.30).

\begin{figure}[ht]
  \begin{tabular*}{0.5\linewidth}{rrr}
\multicolumn{3}{l}{$f(happy, sex)$, see fig. \ref{fig:fact-simple} (right) }\\
  \toprule
  & male & female \\ 
  \midrule
  not too happy & 0.05 & 0.07 \\ 
  pretty happy & 0.25 & 0.31 \\ 
  very happy & 0.14 & 0.18 \\ 
  \bottomrule
  \end{tabular*}
  \\[1em]

  \begin{tabular*}{0.49\linewidth}{rrr}
\multicolumn{3}{l}{$f(sex|happy)$, see fig. \ref{fig:fact-simple} (middle) }\\
  \toprule
  & male & female \\
  \midrule
  not too happy & 0.43 & 0.57 \\
  pretty happy & 0.45 & 0.55 \\ 
  very happy & 0.43 & 0.57 \\
  \bottomrule
  \end{tabular*}
%
\hspace{1em}
  \begin{tabular*}{0.47\linewidth}{rrr}
\multicolumn{3}{l}{$f(happy| sex)$}\\
  \toprule
  & male & female \\ 
  \midrule
  not too happy & 0.12 & 0.12 \\ 
  pretty happy & 0.57 & 0.55 \\ 
  very happy & 0.31 & 0.32 \\
  \bottomrule
  \end{tabular*}
  \\[1em]

  \begin{tabular}{rrrr}
\multicolumn{3}{l}{$f(happy)$, see fig. \ref{fig:fact-simple} (left) and $f(sex)$}\\
  \toprule
  & male & female & \\ 
  \midrule
  not too happy & & & 0.12 \\ 
  pretty happy & & & 0.56 \\ 
  very happy & & & 0.30 \\
  & 0.44 & 0.56 \\ 
  \bottomrule
  \end{tabular}

@@

\begin{figure}[htbp]
  \centering
$f(marital, age) =$\\
\begin{minipage}[t]{0.49\linewidth}
\centering 
$f(marital|age)f(age)$ 
    \includegraphics[width=\linewidth]{part-marital-1}%
 \end{minipage}
\begin{minipage}[t]{0.49\linewidth}
\centering
 $f(age|marital)f(marital)$
    \includegraphics[width=\linewidth]{part-marital-2}
 \end{minipage}
  \caption{(Left) The joint distribution of marital status and age, $f(marital, age)$ as $f(marital|age) \cdot f(age)$, (in decades) partitioned by a vspine and hspine.  (Right) The joint distribution of age and marital status, $f(age, marital)$, partitioned by a vspine and hbar.}
  \label{fig:marital}
\end{figure}

@@

A more theoretical approach in dealing with continuous variables  lies in increasing the number of bins infinitesimally -- this leads from a probability mass function to a density function, making the 1d primitives of hbar and vbar to density plots, and changing the spine-plot representation to a conditional density plot (citation needed - probably Martin's and Heike's paper). While the inclusion of one continuous variable is completely feasible that way, two variables need the use of an additional aesthetic, such a s colour,  to visualize a 2d joint density. Three or more continuous variables break the scope of the product plots framework.

>>>>

We begin in Section~\ref{sec:related} with a review of related work. Then Section~\ref{sec:primitives} motivates the three specific graphical constraints at the heart of product plots: the rectangle. We constrain the rectangle further to produce three 1d atoms (bar, spine, tile) and one 2d atom (the fluct). These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include more visualisations such as histograms, pie charts, cascaded treemaps and weighted plots.

@@

The product plots framework is heavily influenced by the work of Wilkinson \citep{wilkinson:1999, wilkinson:2005}, who suggested that both mosaic plots and treemaps can be described as plots with coordinate systems based on recursive partitioning. Our work builds on this by adding conditioning and defining a ``grammar of area plots'', a domain specific language that describes this small family of plots in more detail. Other similar efforts to build domain specific languages for visualisation are APT \citep{mackinlay:1986} and VisQL \citep{mackinlay:2007}: our framework is driven by similar forces, but carves out a much smaller niche. This makes products plots less expressive, but more efficient, because the number of primitives required to describe the smaller domain is much smaller.

@@

To construct plots of higher-dimensional data sets, we need a way to decompose them into 1d and 2d components. Some statistical vocabulary is useful: one way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}. A {\sc pmf} is a function with $n$ inputs, each indexing one dimension with an integer, which outputs the probability of each combination of inputs. A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all values must sum to one.

@@

Figure~\ref{fig:2d-table} shows three ways to represent the the 2d table of proportions of sex and happiness. The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' -- the bottom left number tells us that these are 0.14. The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex. These correspond to restricting the row or column sums to one, and support questions such as ``what proportion of pretty happy people are female?'' (the table tells us 0.55), or ``what proportion of males are not too happy?'' (the top left number in the table tells us that it's 0.12). Finally, the last table displays the two {\bf marginal distributions} of sex and happiness. These allow us to answer questions like ``what proportion of respondents were male'' (0.44) or ``what proportion of respondents were very happy?'' (0.30).

\begin{figure}[ht]
  \begin{tabular*}{0.5\linewidth}{rrr}
\multicolumn{3}{l}{$f(happy, sex)$}\\
  \toprule
  & male & female \\ 
  \midrule
  not too happy & 0.05 & 0.07 \\ 
  pretty happy & 0.25 & 0.31 \\ 
  very happy & 0.14 & 0.18 \\ 
  \bottomrule
  \end{tabular*}
  \\[1em]

  \begin{tabular*}{0.47\linewidth}{rp{1.5em}p{1.5em}}
\multicolumn{3}{l}{$f(sex|happy)$}\\
  \toprule
  & male & female \\
  \midrule
  not too happy & 0.43 & 0.57 \\
  pretty happy & 0.45 & 0.55 \\ 
  very happy & 0.43 & 0.57 \\
  \bottomrule
  \end{tabular*}
%
\hspace{0.5em}
  \begin{tabular*}{0.47\linewidth}{rp{1.5em}p{1.5em}}
\multicolumn{3}{l}{$f(happy| sex)$}\\
  \toprule
  & male & female \\ 
  \midrule
  not too happy & 0.12 & 0.12 \\ 
  pretty happy & 0.57 & 0.55 \\ 
  very happy & 0.31 & 0.32 \\
  \bottomrule
  \end{tabular*}
  \\[1em]

  \begin{tabular}{rrrr}
\multicolumn{3}{l}{$f(happy)$ and $f(sex)$}\\
  \toprule
  & male & female & \\ 
  \midrule
  not too happy & & & 0.12 \\ 
  pretty happy & & & 0.56 \\ 
  very happy & & & 0.30 \\
  & 0.44 & 0.56 \\ 
  \bottomrule
  \end{tabular}

@@

\begin{figure}[htbp]
  \centering
$f(marital, age) =$\\
\begin{minipage}[t]{0.49\linewidth}
\centering 
$f(marital|age)f(age)$ 
    \includegraphics[width=\linewidth]{part-marital-1}%
 \end{minipage}
\begin{minipage}[t]{0.49\linewidth}
\centering
 $f(age|marital)f(marital)$
    \includegraphics[width=\linewidth]{part-marital-2}
 \end{minipage}
  \caption{(Left) The joint distribution of marital status and age, $f(marital, age)$ as $f(marital|age) \cdot f(age)$, (in decades) partitioned by a vspine and hspine.  (Right) The joint distribution of age and marital status, $f(age, marital)$ as $f(age T|marital) \cdot f(marital)$, partitioned by a vspine and hbar.}
  \label{fig:marital}
\end{figure}

@@

A more theoretical approach in dealing with continuous variables lies in increasing the number of bins infinitesimally, which leads from a probability mass function to a probability density function, turning the 1d primitives bar primitives to density plots, and the 1d spine primitives to a conditional density plot \citep{hofmann:2005}. That approach lends itself to the inclusion of one continuous variable, but two variables need an additional aesthetic, such as colour, to visualize a 2d joint density. Three or more continuous variables break the current scope of the product plots framework.
